Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-04 pnas/Towards.tex
still working on appendix D

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph.
%
\begin{proposition} \label{prop:gamma}
	Let $\sigma > 0$ be fixed. If the mean of $x$ is nonzero, then 
	%
	\begin{align*}
	%
	\gamma & \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	%
	\end{align*}
	%
\end{proposition}
\begin{proof}
In the limit, $(a^1_y)^2=\frac{\gamma^2}{L^2}\sum_{i=0}^{L-1}x[i]x[j]$.  Similarly,  $\sum_{\ell = 1}^{L-1}a_y^2[\ell]=\frac{\gamma}{L}\sum_{\ell = 1}^{L-1}\sum_{i=0}^{L-1-\ell}x[i]x[i+\ell]$ and $a_y^2[0]=\frac{\gamma}{L}\sum_{i=0}^{L-1}x^2[i] + \sigma^2$. The proof is concluded by noting that  $a_x^2[-\ell]=a_x^2[\ell]$. 
%	Plug in the explicit expressions of~\eqref{eq:mean_micrograph} and~\eqref{eq:ac2_micrograph} into the right hand side of the equality.
\end{proof}

@@

\section{Proof of Proposition~\ref{prop:gamma_sigma}} \label{sec:proof_prop_gamma_sigma}
We will prove that both $\sigma$ and $\gamma$ are identifiable from the observed first three moments of $y$. For convenience, we will work with $\beta = \gamma / L$ rather than $\gamma$ itself. We will construct two quadratic equations satisfied by $\beta$ from observed quantities, independent of $\sigma$. Then, we will show that these equations are independent, and hence that $\beta$ is uniquely defined.  Given $\beta$, we can estimate $\sigma$ using Proposition~\ref{prop:gamma}.

@@

Recall that $a_y^1 = \beta(\one^Tx)$ and  
and $a_y^2[0] = \beta\|x\|^2+\sigma^2$, where $\one\in\RL$ stands for vector of ones. Taking the product:
\begin{equation}\label{eq:E1}
\begin{split}
E_1 &:= a_y^1a_y^2[0] =  (\beta(\one^Tx))(\beta\|x\|^2+\sigma^2) \\
& = \sigma^2a_y^1 + L\beta^2F_1,
\end{split}
\end{equation}
where $F_1 := a_x^3[0,0] + \sum_{j=1}^{L-1}(a_x^3[j,j] + a_x^3[0,j])$. 
The terms of $F_1$ can be also estimated from $a_y^3$, while taking the scaling and bias terms into account:
\begin{equation} \label{eq:E2}
E_2:= L\beta F_1 + (2L+1)\sigma^2a_y^1.
\end{equation}
Therefore, from~\eqref{eq:E1} and~\eqref{eq:E2} we get
\begin{equation} \label{eq:E12}
E_2\beta -(2L+1)\sigma^2\beta a_y^1 = E_1-\sigma^2a_y^1.
\end{equation}
Let $E_3:=a_y^2[0] + 2\sum_{j = 1}^{L-1}a_y^2[j]$ and recall from Proposition~\ref{prop:gamma}:
\begin{equation} \label{eq:sigma2}
\sigma^2 = E_3 - (a^1_y)^2/\beta. 
\end{equation} 
Plugging into~\eqref{eq:E12} and rearranging we get 
\begin{equation} \label{eq:quad1}
\mathcal{A}\beta^2 + \mathcal{B}\beta + \mathcal{C} = 0,
\end{equation}
where 
\begin{align*}
\mathcal{A} &= E_2 - (2L+1)a_y^1E_3, \\ 
\mathcal{B} &= -E_1 + (2L+1)(a_y^1)^3 + a_y^1E_3  , \\
\mathcal{C} &= -(a_y^1)^3.
\end{align*}
Importantly, these coefficients are observable quantities.

Next, we derive the second quadratic equation for $\beta$. We notice that 
\begin{equation} \label{eq:E3}
E_4  = \frac{1}{L}(a_y^1)^3 = \frac{1}{L}\beta^3 (\one ^Tx)^3   = \beta^3 F_2,
\end{equation}
where 
\begin{equation*}
F_2 =  a_x^3[0,0] + 3\sum_{j=1}^{L-1}a_x^3[j,j] + 3\sum_{j=1}^{L-1}a_x^3[0,j] + 6\sum_{1\leq i < j\leq L-1}a_x^3[i,j].
\end{equation*}
On the other hand, from $a_y^3$ we can directly estimate $F_2$ up to scale and bias
\begin{equation*} \label{eq:E4}
E_5 = L \beta F_2 + (6L-3)\sigma^2a_y^1.
\end{equation*}
Taking the ratio:
\begin{equation*} 
\frac{E_5}{E_4} = \frac{L}{\beta^2} + \frac{(6L-3)\sigma^2a_y^1}{E_4}, 
\end{equation*}
we conclude:
\begin{equation*}
\sigma^2 = \frac{E_5}{a_y^1(6L-3)}  - \frac{LE_4}{\beta^2a_y^1(6L-3)}.
\end{equation*}
Using~\eqref{eq:sigma2} and rearranging we get the second quadratic:
\begin{equation} \label{eq:quad2}
\mathcal{D}\beta^2 + \mathcal{E}\beta + \mathcal{F} = 0,
\end{equation}
where
\begin{align*}
\mathcal{D} &= E_3 - \frac{E_5}{a_y^1(6L-3)}, \\ 
\mathcal{E} &= -(a_y^1)^2, \\
\mathcal{F} &= \frac{LE_4}{a_y^1(6L-3)}.
\end{align*}

>>>>

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph.
%
\begin{proposition} \label{prop:gamma}
	Let $\sigma > 0$ be fixed. If the mean of $x$ is nonzero, then 
	%
	\begin{align*}
	%
	\gamma & \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	%
	\end{align*}
	%
\end{proposition}
\begin{proof}
In the limit, $(a^1_y)^2=\frac{\gamma^2}{L^2}\sum_{i=0}^{L-1}x[i]x[j]$. \TODO{In the previous equation, index $j$ is unaccounted for} Similarly,  $\sum_{\ell = 1}^{L-1}a_y^2[\ell]=\frac{\gamma}{L}\sum_{\ell = 1}^{L-1}\sum_{i=0}^{L-1-\ell}x[i]x[i+\ell]$ and $a_y^2[0]=\frac{\gamma}{L}\sum_{i=0}^{L-1}x^2[i] + \sigma^2$. The proof is concluded by noting that  $a_x^2[-\ell]=a_x^2[\ell]$. 
%	Plug in the explicit expressions of~\eqref{eq:mean_micrograph} and~\eqref{eq:ac2_micrograph} into the right hand side of the equality.
\end{proof}

@@

\section{Proof of Proposition~\ref{prop:gamma_sigma}} \label{sec:proof_prop_gamma_sigma}

\TODO{Nicolas: I'm pretty sure this proof assumes $L \geq 3$: let's add it as an assumption in the prop.}

We will prove that both $\sigma$ and $\gamma$ are identifiable from the observed first three moments of $y$. For convenience, we will work with $\beta = \gamma / L$ rather than $\gamma$ itself. We will construct two quadratic equations satisfied by $\beta$ from observed quantities, \TODO{independent of $\sigma$---they're not independent of $\sigma$; did you mean `independently'?}. Then, we will show that these equations are independent, and hence that $\beta$ is uniquely defined.  Given $\beta$, we can estimate $\sigma$ using Proposition~\ref{prop:gamma}.

@@

Recall that $a_y^1 = \beta(\one^Tx)$ and $a_y^2[0] = \beta\|x\|^2+\sigma^2$, \TODO{the latter statements are only true in expectation, or almost surely in the limit though; should revise whole proof accordingly} where $\one\in\RL$ is the vector of all-ones. Taking the product:
\begin{equation}\label{eq:E1}
\begin{split}
E_1 &:= a_y^1a_y^2[0] =  (\beta(\one^Tx))(\beta\|x\|^2+\sigma^2) \\
& = \sigma^2a_y^1 + L\beta^2F_1,
\end{split}
\end{equation}
where $F_1 := a_x^3[0,0] + \sum_{j=1}^{L-1}(a_x^3[j,j] + a_x^3[0,j])$. 
% Nicolas: the expression for F_1 is correct, see identifiability_of_gamma_and_sigma_Sep4_2018.m and notes Sep 4, 2018
The terms of $F_1$ can be also estimated from $a_y^3$, while taking the scaling and bias terms into account. This yields another observable:
\begin{align} 
E_2 & := a_y^3[0,0] + \sum_{j=1}^{L-1}(a_y^3[j,j] + a_y^3[0,j]) \nonumber\\
	& = L\beta F_1 + (2L+1)\sigma^2a_y^1. \label{eq:E2}
\end{align}
% Nicolas: checked this one too.
Therefore, from~\eqref{eq:E1} and~\eqref{eq:E2} we get
\begin{equation} \label{eq:E12}
E_2\beta -(2L+1)\sigma^2\beta a_y^1 = E_1-\sigma^2a_y^1.
\end{equation}
Let $E_3:=a_y^2[0] + 2\sum_{j = 1}^{L-1}a_y^2[j]$ and recall from Proposition~\ref{prop:gamma}:
\begin{equation} \label{eq:sigma2}
\sigma^2 = E_3 - (a^1_y)^2/\beta. 
\end{equation} 
Plugging into~\eqref{eq:E12} and rearranging we get a first quadratic equation in $\beta$,
\begin{equation} \label{eq:quad1}
\mathcal{A}\beta^2 + \mathcal{B}\beta + \mathcal{C} = 0,
\end{equation}
where 
\begin{align*}
\mathcal{A} &= E_2 - (2L+1)a_y^1E_3, \\ 
\mathcal{B} &= -E_1 + (2L+1)(a_y^1)^3 + a_y^1E_3  , \\
\mathcal{C} &= -(a_y^1)^3.
\end{align*}
Importantly, these coefficients are observable quantities. 
% Nicolas, Sep 4 2018: everything checked and correct until here.

Next, we derive the second quadratic equation for $\beta$. We notice that 
\begin{equation} \label{eq:E3}
E_4 := \frac{1}{L}(a_y^1)^3 = \frac{1}{L}\beta^3 (\one ^Tx)^3   = \beta^3 F_2,
\end{equation}
where 
\begin{equation*}
F_2 =  a_x^3[0,0] + 3\sum_{j=1}^{L-1}a_x^3[j,j] + 3\sum_{j=1}^{L-1}a_x^3[0,j] + 6\sum_{1\leq i < j\leq L-1}a_x^3[i,j].
\end{equation*}
On the other hand, from $a_y^3$ we can directly estimate $F_2$ up to scale and bias
\begin{equation*} \label{eq:E4}
E_5 := L \beta F_2 + (6L-3)\sigma^2a_y^1.
\end{equation*}
Taking the ratio:
\begin{equation*} 
\frac{E_5}{E_4} = \frac{L}{\beta^2} + \frac{(6L-3)\sigma^2a_y^1}{E_4}, 
\end{equation*}
we conclude:
\begin{equation*}
\sigma^2 = \frac{E_5}{a_y^1(6L-3)}  - \frac{LE_4}{\beta^2a_y^1(6L-3)}.
\end{equation*}
Using~\eqref{eq:sigma2} and rearranging we get the second quadratic:
\begin{equation} \label{eq:quad2}
\mathcal{D}\beta^2 + \mathcal{E}\beta + \mathcal{F} = 0,
\end{equation}
where
\begin{align*}
\mathcal{D} &= E_3 - \frac{E_5}{a_y^1(6L-3)}, \\ 
\mathcal{E} &= -(a_y^1)^2, \\
\mathcal{F} &= \frac{LE_4}{a_y^1(6L-3)}.
\end{align*}
